---
title: 'Reproducible Research: PA_1'
author: "David Anderson"
date: "March 5, 2016"
output: html_document
---

This is a file that details the steps to complete an analysis of fitness tracker activity.

This exercise requires knitr to create the output and we will use ggplot2 to generate visualizations.

```{r}
library(knitr, quiet=T)
library(plyr, quiet=T)  
library(ggplot2, quiet=T)
opts_chunk$set(dev="png", fig.path="figure/")
```

Read in the data to the fitivity variable:

```{r, echo=FALSE}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
fitivity <- read.csv('activity.csv', header="TRUE",colClasses=c("numeric", "character", "numeric"),
                     stringsAsFactors=FALSE)
```

Cleaning up the data a bit to ensure that date information is easy to manipulate:

```{r}
fitivity$date.factor <- as.factor(fitivity$date)
fitivity$date <- as.Date(fitivity$date, format="%Y-%m-%d")
``` 

1.  Steps per day histogram, ignore NA

```{r}
fitivity.daily <- ddply(activity, .(date.factor), summarize,
                        total.steps=sum(steps, na.rm=TRUE))
hist(fitivity.daily$total.steps, col="blue", main="Fitness Activity",
     xlab="Daily Steps")
```

!(figure/histogram-1.png)

The mean and median per interval are:

```r
mn <- mean(fitivity$steps, na.rm=TRUE)
md <- median(fitivity$steps, na.rm=TRUE)
cat(paste("The mean number of steps per interval is", 
          format(mn, digits=4), "and the median is", 
          format(md, digits=2), "(ignoring missing values)"))
```

```
## The mean number of steps per interval is 37.38 and the median is 0 (ignoring missing values)
```

And per day (which is the question from the assignment)

```r
mn <- mean(fitivity.daily$total.steps, na.rm=TRUE)
md <- median(fitivity.daily$total.steps, na.rm=TRUE)
cat(paste("The mean number of steps per day is", 
          format(mn, digits=4), "and the median is", 
          format(md, digits=2), "(ignoring missing values"))
```

```
## The mean number of steps per day is 9354 and the median is 10395 (ignoring missing values
```

## What is the average daily activity pattern?

Here we will take advantage of the fact that the interval variable
*wraps* at the end of each day.  Again, we are ignoring NA values.


```r
fitivity.pattern <- ddply(fitivity, "interval", summarize,
                          mean=mean(steps, na.rm=TRUE))
g <- ggplot(fitivity.pattern, aes(interval, mean))
g + geom_line(color="blue") + 
  labs(title="Fitness tracker", 
       x="Interval throughout day (min)",
       y="Mean number of steps")
```

!(figure/activitypattern-1.png) 

What interval has the highest average number of steps?

```r
max.interval <- fitivity.pattern$interval[which.max(fitivity.pattern$mean)]
cat(paste("Interval corresponding to maximum average number of steps:",
          max.interval,"minutes"))
```

```
## Interval corresponding to maximum average number of steps: 835 minutes
```

## Imputing missing values

How many rows have missing values?  Note that this is not the
same as the number of missing elements if there are more
than two NAs in a row.

```r
nmissing <- sum(!complete.cases(fitivity))
cat(paste("Number of rows with missing data:", nmissing))
```

```
## Number of rows with missing data: 2304
```

We fill missing values for the number of steps with the mean number 
for that interval.   Note that this will make the steps non-integral.


```r
impute.value <- function(steps, interval) {
    if (is.na(steps)) {
        fitivity.pattern[fitivity.pattern$interval==interval,]$mean
    } else {
        steps
    }
}
imputed.fitivity <- fitivity
imputed.fitivity$steps <- mapply(impute.value, fitivity$steps, 
                                fitivity$interval)
```

Now see how different the mean and median are for the total
number per day with and without imputation:

```r
total.steps <- tapply(fitivity$steps, fitivity$date.factor,
                      sum, na.rm=TRUE)
total.steps.imputed <- tapply(imputed.fitivity$steps,
                              imputed.fitivity$date.factor, sum)
cat(paste("For the raw data the mean and median per day are:",
          format(mean(total.steps)), "and", median(total.steps)))
```

```
## For the raw data the mean and median per day are: 9354.23 and 10395
```

```r
cat(paste("For the imputed the mean and median per day are:",
          format(mean(total.steps.imputed)), "and", 
          format(median(total.steps.imputed))))
```

```
## For the imputed the mean and median per day are: 10766.19 and 10766.19
```
In this case imputation increases both the mean and median.  Note that they
are not exactly equal after imputation if more digits are shown, but are 
quite close.  Imputing using the median would have a very different 
effect, and would lower the mean.

Make a histogram of the mean number per day after imputation.

```r
hist(total.steps.imputed, col="blue", main="Fitness Tracker Activity",
     xlab="Number of Steps Per Day (imputed)")
```

!(figure/meanperdayimpute-1.png) 

Check to make sure we filled in all the missing values:

```r
nmissing <- sum(!complete.cases(imputed.fitivity))
cat(paste("After imputation, number of rows with missing data:", nmissing))
```

```
## After imputation, number of rows with missing data: 0
```

## Are there differences in activity patterns between weekdays and weekends?

Add a factor variable for weekday vs. weekend:

```r
daytype <- function(date) 
    if (weekdays(date) %in% c("Saturday", "Sunday")) "weekend" else "weekday"
imputed.fitivity$day.type <- as.factor(sapply(imputed.fitivity$date, daytype))
```

And look at the mean number of steps per day per date type:

```r
steps.day.daytype <- ddply(imputed.fitivity, .(interval, day.type),
                           summarize, steps=mean(steps))
ggplot(steps.day.daytype, aes(interval, steps)) + 
    geom_line() + facet_grid(day.type ~ .) +
    labs(x="Time of Day (min)", y="Number of steps",
         title="Activity patterns on weekdays vs. weekends")
```

!(figure/mean.per.day.by.daytype-1.png) 
